---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# spatialData

<!-- badges: start -->
[![R-CMD-check](https://github.com/BlasBenito/spatialData/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/BlasBenito/spatialData/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`spatialData` is a data-only R package providing ready-to-use spatial datasets for ecological modeling. It was originally developed to supply example data for the R packages [`collinear`](https://blasbenito.github.io/collinear/) and [`spatialRF`](https://blasbenito.github.io/spatialRF/), but the datasets are equally suited for teaching spatial analysis and species distribution modeling. Each dataset comes pre-processed with response variables and environmental predictors, so you can jump straight into the analysis without spending time on data preparation.

## Installation

Install the development version from [GitHub](https://github.com/blasbenito/spatialData):

```r
# Option 1: pak (recommended)
# install.packages("pak")
pak::pkg_install("blasbenito/spatialData")

# Option 2: remotes
# install.packages("remotes")
remotes::install_github("blasbenito/spatialData")
```

## Datasets

All datasets are loaded lazily via `data()`. Most are `sf` data frames with point or polygon geometries; `interaction` is a plain data frame.

| Dataset | Description | Spatial scope | Temporal scope | Rows | Responses | Predictors | CRS | Geometry |
|:--------|:------------|:--------------|:---------------|-----:|----------:|-----------:|:----|:---------|
| [`andalusia`](https://blasbenito.github.io/spatialData/articles/andalusia.html) | Plant presences and and environmental predictors | Andalusia, Spain | 2000 - 2010 | 8,666 | 2 | 20 | EPSG:25830 | POINT |
| [`neanderthal`](https://blasbenito.github.io/spatialData/articles/neanderthal.html) | Neanderthal presence/absence and paleoclimate predictors | Europe | ~120,000 BP (MIS 5e) | 245 | 1 | 25 | EPSG:4326 | POINT |
| [`quercus`](https://blasbenito.github.io/spatialData/articles/quercus.html) | Distribution of European oaks  and bioglimatic predictors | Europe | present | 6,728 | 1 | 31 | EPSG:4326 | POINT |
| [`vi`](https://blasbenito.github.io/spatialData/articles/vi.html) | NDVI records with environmental predictors | Global | 1999–2019 | 9,265 | 5 | 58 | EPSG:4326 | POINT |
| [`plantae`](https://blasbenito.github.io/spatialData/articles/plantae.html) | Plant diversity of the global ecoregions | Global | 2020 | 662 | 53 | 84 | EPSG:4326 | POINT (centroids) |
| [`trees`](https://blasbenito.github.io/spatialData/articles/trees.html) | Tree species richness and environmental predictors | Americas | 2012 | 3,373 | 1 | 50 | EPSG:4326 | POLYGON |
| [`communities`](https://blasbenito.github.io/spatialData/articles/communities.html) | Plant community presence and climate predictors | Sierra Nevada, SE Spain | 2009 | 7,300 | 6 | 9 | EPSG:25830 | POINT |
| [`interaction`](https://blasbenito.github.io/spatialData/articles/interaction.html) | Butterfly and host plant presence and environmental predictors | Sierra Nevada, SE Spain | 2014 | 1000 | 3 | 10 | EPSG:25830 | POINT |
| [`linaria`](https://blasbenito.github.io/spatialData/articles/linaria.html) | *Linaria nigricans* and greenhouse records with environmental predictors | Eastern Andalusia, Spain | | 7,386 | 2 | 20 | EPSG:25830 | POINT |

## Extra Data

Several datasets ship with companion download functions that retrieve larger or
complementary files from [spatialDataExtra](https://github.com/BlasBenito/spatialDataExtra).

| Dataset | Function | Description | Format | Geometry / Bands | Resolution | CRS |
|:--------|:---------|:------------|:-------|:-----------------|:-----------|:----|
| [`neanderthal`](https://blasbenito.github.io/spatialData/articles/neanderthal.html) | `neanderthal_extra()` | 25-band palaeoclimate raster (bioclim + topography) | GeoTIFF | 25 bands | 0.18° | EPSG:4326 |
| [`quercus`](https://blasbenito.github.io/spatialData/articles/quercus.html) | `quercus_extra()` | 31-band bioclimatic/environmental raster | GeoTIFF | 31 bands | 0.167° | EPSG:4326 |
| [`vi`](https://blasbenito.github.io/spatialData/articles/vi.html) | `vi_extra()` | Extended NDVI point dataset (30,000 records) | GeoPackage | POINT | 30,000 rows | EPSG:4326 |
| [`plantae`](https://blasbenito.github.io/spatialData/articles/plantae.html) | `plantae_extra()` | Ecoregion polygons with full diversity data | GeoPackage | MULTIPOLYGON | 662 features | EPSG:4326 |
| [`trees`](https://blasbenito.github.io/spatialData/articles/trees.html) | `trees_extra()` | Individual tree species presence points | GeoPackage | POINT | 523,308 rows | EPSG:4326 |
| [`communities`](https://blasbenito.github.io/spatialData/articles/communities.html) | `communities_extra_2010()` | Baseline (2010) climate raster for Sierra Nevada | GeoTIFF | 9 bands | 100 m | EPSG:25830 |
| [`communities`](https://blasbenito.github.io/spatialData/articles/communities.html) | `communities_extra_2050()` | Projected (2050) climate raster for Sierra Nevada | GeoTIFF | 9 bands | 100 m | EPSG:25830 |
| [`communities`](https://blasbenito.github.io/spatialData/articles/communities.html) | `communities_extra_2100()` | Projected (2100) climate raster for Sierra Nevada | GeoTIFF | 9 bands | 100 m | EPSG:25830 |
| [`interaction`](https://blasbenito.github.io/spatialData/articles/interaction.html) | `interaction_extra()` | 10-band environmental raster for Sierra Nevada | GeoTIFF | 10 bands | 100 m | EPSG:25830 |
| [`linaria`](https://blasbenito.github.io/spatialData/articles/linaria.html) | `linaria_extra()` | 20-band environmental raster for Eastern Andalusia | GeoTIFF | 20 bands | 400 m | EPSG:25830 |
| [`andalusia`](https://blasbenito.github.io/spatialData/articles/andalusia.html) | `andalusia_extra()` | 20-band environmental raster for Andalusia | GeoTIFF | 20 bands | 400 m | EPSG:25830 |

## Citation

The license of `spatialData` requires the following citation:

```{r}
citation("spatialData")
```


## License

CC BY 4.0 — see [LICENSE.md](LICENSE.md) for details.
