---
title: "andalusia"
output:
  rmarkdown::html_document:
    toc: true
    toc_title: "Contents"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval      = TRUE,
  collapse  = TRUE,
  comment   = "#>",
  out.width = "100%"
)
```

## Overview

The `andalusia` dataset is an `sf` data frame with coordinate reference system ETRS89 / UTM zone 30N (EPSG:25830), containing 37,773 presence records for 90 plant species and 8,692 background points (46,465 rows total) from Andalusia, Spain. The data frame has only 3 columns: `species` (character), `presence` (integer), and `geometry` (POINT).

Environmental predictors are **not** columns in `andalusia` — they live in the companion 20-layer raster downloaded via `andalusia_extra()`, and their names are stored in `andalusia_predictors`.

This dataset is well suited for multi-species SDMs, multiclass classification, and spatial richness mapping.

## Setup

The code below loads the `R` packages and example data required for this tutorial.

```{r, include = FALSE}
if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak")
}
pak::pkg_install(
  c(
    "sf",
    "dplyr",
    "terra",
    "stars",
    "collinear",
    "ranger",
    "mapview",
    "viridis"
  ),
  ask = FALSE
)
```

```{r, message = FALSE, warning = FALSE, echo = FALSE}
library(dplyr)
library(terra)
library(sf)
library(collinear)
library(ranger)
library(mapview)
library(viridis)
library(spatialData)

data(
  andalusia,
  andalusia_responses,
  andalusia_predictors,
  package = "spatialData"
)
```

## Data Structure

The dataset is an `sf` data frame with `r nrow(andalusia)` rows and `r ncol(andalusia)` columns.

```{r}
dplyr::glimpse(andalusia)
```

The `species` column identifies each record as a named species or `"background"`. The table below shows the number of records per species or group.

```{r}
andalusia |>
  sf::st_drop_geometry() |>
  dplyr::group_by(species) |>
  dplyr::summarise(n = dplyr::n()) |>
  dplyr::arrange(dplyr::desc(n))
```

Background points were generated randomly across the entire study area to sample the range of environmental conditions available.

The map below shows all species presences. Background points were excluded for clarity.

```{r map, message = FALSE}
mapview::mapview(
  andalusia |>
    dplyr::filter(
      species != "background"
    ),
  zcol = "species",
  layer.name = "Species",
  col.regions = viridis::turbo(n = 90),
  legend = FALSE
)
```

The environmental raster matching this dataset is downloaded via `andalusia_extra()`, which returns a multilayer `SpatRaster` object (R package [`terra`](https://cran.r-project.org/package=terra)).

```{r load_raster, message = FALSE}
andalusia_env <- spatialData::andalusia_extra()
plot(andalusia_env)
```

The `SpatRaster` object contains `r length(andalusia_predictors)` layers, all named in the vector `andalusia_predictors`.

```{r}
andalusia_predictors
```

Each cell of the raster dataset represents 400 × 400 meters on the ground.

```{r}
terra::res(andalusia_env)
```

To pair records in `andalusia` with the cell values in `andalusia_env`, we use `terra::extract()`.


```{r}
env_df <- terra::extract(
  x = andalusia_env,
  y = andalusia,
  ID = FALSE
)

andalusia <- dplyr::bind_cols(
  andalusia,
  env_df
) |> 
  na.omit()

dplyr::glimpse(andalusia)
```

To speed up code execution in the next section, we are going to reduce the resolution of `andalusia_env` by a factor of 4 via aggregation.

```{r}
andalusia_env <- terra::aggregate(
  x = andalusia_env,
  fact = 4,
  fun = mean
)

res(andalusia_env)
```

## Example Usage

This section demonstrates a complete multi-species distribution modelling workflow:

- **Species Distribution Modelling**: fit a `ranger::ranger()` binary random forest for *Pinus halepensis*, then loop over all species.
- **Species Richness Map**: sum binary predictions to produce a species richness raster.

Before fitting models, we convert `presence` to a factor (required by `ranger` for classification) and split the data into two objects: `species` (presence records) and `background`.

```{r}
andalusia <- andalusia |>
  dplyr::mutate(
    presence = factor(presence)
  )

background <- andalusia |>
  dplyr::filter(
    species == "background"
  )

species <- andalusia |>
  dplyr::filter(
    species != "background"
  )
```




### Modelling steps for one species

This section shows the modelling steps for a single species.

```{r}
species_name <- "Pinus_halepensis"
```

First, we filter the records belonging to the target species.

```{r}
species_data <- species |>
  dplyr::filter(
    species == species_name
    )
```

Next, we build a 10 km buffer around the species presences to exclude background points that fall within it — records too close to presences would dilute the contrast the model needs to learn habitat boundaries.

```{r}
species_data_buffer <- sf::st_buffer(
  x = species_data,
  dist = 10000
) |> 
  sf::st_union()
```


We then identify which background points fall outside the buffer using `sf::st_contains()` and retain only those.

```{r}
background_outside_buffer <- !sf::st_contains(
  x = species_data_buffer,
  y = background,
  sparse = FALSE
) |>
  as.vector()

species_background <- background[background_outside_buffer, ]
```

The map below shows the 10 km buffer around the species presences and the background points that remain after filtering.

```{r}
mapview(species_data_buffer, col.regions = "gray50") + mapview(species_background, col.regions = "red4")
```

We combine the species presences and the filtered background into one training data frame.

```{r}
species_data <- dplyr::bind_rows(
  species_data,
  species_background
)
```


We train a binary random forest using all predictors and class weights to account for the imbalance between presences and background records. `num.trees = 100` and `min.node.size = 30` are set below their defaults (500 and 5, respectively) to reduce computation time — increase them for production models.

```{r}
species_model <- ranger::ranger(
  formula = presence ~ .,
  data = sf::st_drop_geometry(species_data)[, c("presence", andalusia_predictors)],
  case.weights = collinear::case_weights(
    x = species_data$presence
  ),
  num.trees = 100,
  min.node.size = 30
)
```

We then apply the fitted model to every cell of `andalusia_env` to produce a binary habitat suitability raster.

```{r}
species_prediction <- terra::predict(
  object = andalusia_env,
  model = species_model
)
```

The prediction is a binary map.

```{r}
mapview(
  species_data |> 
    dplyr::filter(
      species == species_name
    ),
  layer.name  = paste0("Presence ", species_name),
  alpha.regions = 0,
  color       = "black",
  col.regions = "black"
) +
  mapview(
    species_prediction,
    layer.name = paste0("Prediction ", species_name),
    maxpixels  = terra::ncell(species_prediction),
    na.alpha   = 0
  )
```


### Modelling all species

We extract the names of all 90 species to iterate over them.

```{r}
species_names <- andalusia |>
  dplyr::filter(species != "background") |>
  dplyr::distinct(species) |>
  dplyr::pull(species)

species_names
```

We now apply the same modelling pipeline to all 90 species. The result is a list of per-species binary rasters, stacked into a single `SpatRaster` with `terra::rast()`.

```{r}
species_predictions <- lapply(
  X = species_names,
  FUN = function(species_name) {

    species_data <- species |>
      dplyr::filter(species == species_name)

    species_data_buffer <- sf::st_buffer(
      x = species_data,
      dist = 10000
    ) |>
      sf::st_union()

    background_outside_buffer <- !sf::st_contains(
      x = species_data_buffer,
      y = background,
      sparse = FALSE
    ) |>
      as.vector()

    species_background <- background[background_outside_buffer, ]

    species_data <- dplyr::bind_rows(
      species_data,
      species_background
    )

    species_model <- ranger::ranger(
      formula = presence ~ .,
      data = sf::st_drop_geometry(species_data)[, c("presence", andalusia_predictors)],
      case.weights = collinear::case_weights(
        x = species_data$presence
      ),
      num.trees     = 100,
      min.node.size = 30
    )

    terra::predict(
      object = andalusia_env,
      model = species_model
    )
  }
) |>
  terra::rast()
```

### Species Richness Map

Because each model is a binary classifier, every per-species prediction is already a 0/1 raster: 1 where the model predicts suitable habitat, 0 elsewhere. Summing these layers gives a species richness raster where each pixel counts how many of the 90 species have predicted suitable habitat at that location.

```{r}
species_predictions_sum <- sum(species_predictions)

mapview(
  species_predictions_sum,
  col.regions = viridis::turbo(n = 100)
)
```

## Conclusion

This vignette demonstrated a complete multi-species modelling workflow using the `andalusia` dataset. After extracting environmental predictors from the companion raster via `terra::extract()`, we fitted `ranger::ranger()` binary random forest models for all 90 plant species using background points outside a 10 km buffer around each species' presences, and summed the per-species binary predictions into a species richness raster where each pixel value counts how many species have predicted suitable habitat at that location. Areas of high predicted richness correspond to climatically and topographically heterogeneous zones in Andalusia. Beyond the SDM workflow shown here, the dataset supports multiclass classification (using the `species` column), habitat connectivity analysis, and spatial conservation prioritisation.

Happy modelling!

## References

Benito, B.M., Lorite, J., Pérez-Pérez, R., Gómez-Aparicio, L., & Peñas, J. (2014). Forecasting plant range collapse in a mediterranean hotspot: when dispersal uncertainties matter. *Diversity and Distributions*, 20(1), 72–83. https://doi.org/10.1111/ddi.12148

```{r, echo = FALSE}
unlink("andalusia_env.tif")
```
