---
title: "interaction"
output:
  rmarkdown::html_document:
    toc: true
    toc_title: "Contents"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval      = TRUE,
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.align = "center",
  dpi = 150,
  out.width = "100%"
)
```

## Overview

Species distribution models (SDMs) typically rely on abiotic predictors — climate, topography, remote sensing — but many species are shaped just as strongly by who they live with. Obligate specialists, in particular, cannot persist where their host is absent, no matter how suitable the climate. A sequential modelling approach captures this by first modelling the host species and then feeding those predictions as a biotic predictor into the dependent species' model.

The `interaction` dataset is designed for exactly this workflow. It contains occurrence records for two Sierra Nevada (SE Spain) species: the host plant *Androsace vitaliana* (`host_plant`) and the obligate specialist butterfly *Agriades zullichi* (`butterfly`), together with random background points.

The dataset includes 3 response variables and 10 environmental predictors derived from remote sensing, climate, and topographic data at 100 m resolution.

## Setup

The code below loads the R libraries and example data required to run this tutorial.

```{r}
if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak")
}
pak::pkg_install(
  c(
    "dplyr",
    "terra",
    "sf",
    "stars",
    "mapview",
    "collinear",
    "spatialRF"
  ),
  ask = FALSE
)
```

```{r, message = FALSE, warning = FALSE, echo = FALSE}
library(dplyr)
library(terra)
library(sf)
library(mapview)
library(collinear)
library(spatialRF)
library(spatialData)

data(
  interaction,
  interaction_responses,
  interaction_predictors,
  package = "spatialData"
)
```



## Data Structure

The dataset `interaction` is an sf data frame with `r nrow(interaction)` rows and `r ncol(interaction)` columns.

```{r}
dplyr::glimpse(interaction)
```

The three response variables are:

- `class`: factor with levels `"butterfly"`, `"host_plant"`, and `"background"`, identifying the observation type of each row. You can interpret background points as the presence of "anything that is not butterfly or host plant".

```{r}
table(interaction$class)
```

The map below shows a spatial representation of this column.

```{r}
mapview::mapview(interaction, zcol = "class")
```


- `butterfly`: integer with values 1 and 0 representing butterfly and background records, and `NA` for the host plant records (to simplify subsetting).

```{r}
table(na.omit(interaction$butterfly))
```


- `host_plant`: integer with values 1 and 0 indicating host plant presence and background records, and `NA` for the butterfly records.


```{r}
table(na.omit(interaction$host_plant))
```

The dataset includes 10 environmental predictors covering remote sensing, climate, and topography at 100m resolution.

```{r}
interaction_predictors
```

## Example Usage

In this example we train two models:

- *Host plant model*: Random forest model of the host plant using abiotic predictors.

- *Butterfly model*: Random forest model of the butterfly using the prediction of the host plant model as a biotic predictor along with the abiotic ones.

### Host plant model

In this section we train and analyze a habitat suitability model for the host plant based on the abiotic predictors.

#### Data Preparation

To train the host plant model we first prepare a training data frame with the plant and background records, and a data frame with case coordinates to be used in spatial cross-validation.

```{r}
host_plant_df <- interaction |> 
  dplyr::select(
    -butterfly
  ) |> 
  na.omit()

host_plant_xy <- host_plant_df |> 
    sf::st_coordinates() |> 
    as.data.frame() |> 
    dplyr::rename(
      x = X,
      y = Y
    )
```


#### Multicollinearity Filtering

The function [`collinear::collinear()`](https://blasbenito.github.io/collinear/reference/collinear.html) selects a non-redundant subset of the most relevant predictors. The function ranks all predictors by the AUC of univariate binomial random forest models with case weights, and then applies a multicollinearity filtering that preserves the most important ones.

```{r}
host_plant_selection <- collinear::collinear(
  df = host_plant_df,
  responses = "host_plant",
  predictors = interaction_predictors,
  f = collinear::f_binomial_rf,
  quiet = TRUE
)
```

Here are the selected predictors for the host plant:

```{r}
host_plant_selection$host_plant$selection
```

#### Model Training

Here we train a random forest model with `spatialRF::rf()`, which computes case weights internally to control for class imbalance.

```{r, fig.width=6, fig.height=6}
host_plant_model <- spatialRF::rf(
  data = sf::st_drop_geometry(host_plant_df),
  dependent.variable.name = "host_plant",
  predictor.variable.names = host_plant_selection$host_plant$selection,
  verbose = FALSE
)
```

To understand how well the model works on unseen data, we perform spatial cross-validation via `spatialRF::rf_evaluate()`, which partitions the data into spatially disjunct training and testing folds, providing an honest assessment of the model's ability to predict habitat suitability for the host plant.

```{r, fig.width=4, fig.height=3}
host_plant_model <- spatialRF::rf_evaluate(
  model = host_plant_model,
  xy = host_plant_xy,
  metrics = "auc",
  verbose = FALSE
)

host_plant_model$evaluation$aggregated |> 
  dplyr::filter(model == "Testing") |> 
  dplyr::glimpse()
```

The dataframe shows the stats of the AUC scores on 30 cross-validation repetitions. The median score indicates that the model has no issues discriminating between host plant and background records.

Finally, we predict the host plant suitability for all rows in the dataframe `interaction`, and write the result to the new column `host_plant_suitability`. 

```{r}
interaction$host_plant_suitability <- predict(
  host_plant_model,
  data = sf::st_drop_geometry(interaction)
)$predictions
```


### Butterfly Model

Now that we have a trained and evaluated host plant model, let's use its predictions as a biotic predictor for the butterfly.

#### Data Preparation

We prepare `butterfly_df` and `butterfly_xy` in the same way we did for the host plant. Additionally, we add the new biotic predictor `host_plant_suitability` to the vector `interaction_predictors`.


```{r}
butterfly_df <- interaction |> 
  dplyr::select(
    -host_plant
  ) |> 
  na.omit()

butterfly_xy <- butterfly_df |> 
    sf::st_coordinates() |> 
    as.data.frame() |> 
    dplyr::rename(
      x = X,
      y = Y
    )

butterfly_predictors <- c(
  interaction_predictors, 
  "host_plant_suitability"
  )
```

#### Multicollinearity Filtering

Here we use `collinear::collinear()` again to select the best subset of non-collinear predictors for the butterfly.
 
```{r}
butterfly_selection <- collinear::collinear(
  df = butterfly_df,
  responses = "butterfly",
  predictors = butterfly_predictors,
  f = collinear::f_binomial_rf,
  quiet = TRUE
)
```

The selected non-collinear predictors include the host plant's suitability.

```{r}
butterfly_selection$butterfly$selection
```

#### Model Training

We use the same modelling pattern we used with the host plant to model the suitability of the butterfly and evaluate the model via cross-validation.

```{r}
butterfly_model <- spatialRF::rf(
  data = sf::st_drop_geometry(butterfly_df),
  dependent.variable.name = "butterfly",
  predictor.variable.names = butterfly_selection$butterfly$selection,
  verbose = FALSE
)

butterfly_model <- spatialRF::rf_evaluate(
  model = butterfly_model,
  xy = butterfly_xy,
  metrics = "auc",
  verbose = FALSE
)

butterfly_model$evaluation$aggregated |> 
  dplyr::filter(model == "Testing") |> 
  dplyr::glimpse()
```
The cross-validation shows that the model has an almost perfect ability to discriminate between butterfly and background records.

```{r, fig.width=4.5, fig.height=3.5}
spatialRF::plot_importance(butterfly_model)
```

The permutation importance scores plotted above indicate that `host_plant_suitability` contributes most to model error when permuted.

```{r, fig.width=5.5, fig.height=3.5}
spatialRF::plot_response_curves(
  model = butterfly_model,
  variables = "host_plant_suitability"
  )
```

The response curves show a clear sigmoidal increase in butterfly suitability as host plant suitability increases, with little variation across the quantiles of the remaining predictors — consistent with an obligate dependency.

### Habitat Suitability Maps

With both models trained, we can now project habitat suitability across the Sierra Nevada. The package `spatialData` provides the raster dataset `sierra_nevada_env` for exactly this purpose.

```{r}
sierra_nevada_env <- spatialData::interaction_extra()
class(sierra_nevada_env)
names(sierra_nevada_env)
```

We can project the model onto this raster dataset using `terra::predict()`.

```{r}
host_plant_prediction <- terra::predict(
  object = sierra_nevada_env,
  model = host_plant_model,
  na.rm = TRUE
)
names(host_plant_prediction) <- "host_plant_suitability"

mapview(
  host_plant_prediction, 
  layer.name = "Host plant suitability",
  na.color = "transparent"
  )
```

To predict the habitat suitability for the butterfly, we first need to add the raster prediction of the host plant to `sierra_nevada_env`, and then apply `terra::predict()`.

```{r}
sierra_nevada_env <- c(
  sierra_nevada_env, 
  host_plant_prediction
  )

butterfly_prediction <- terra::predict(
  object = sierra_nevada_env,
  model = butterfly_model,
  na.rm = TRUE
)

names(butterfly_prediction) <- "butterfly_suitability"

mapview(
  butterfly_prediction, 
  layer.name = "Butterfly suitability",
  na.color = "transparent"
  )
```
The habitat suitability maps of both species look very similar, but there are actual differences between them. The code below generates a map of differences between predicted suitabilities.

```{r}
#raster of prediction differences
difference_predictions <- butterfly_prediction - host_plant_prediction

#compute maximum absolute value
difference_predictions_max <- terra::global(
  x = difference_predictions, 
  fun = range, 
  na.rm = TRUE
  ) |> 
  abs() |> 
  max()

#compute breaks for color palette
color_breaks <- seq(
  from = -difference_predictions_max, 
  to = difference_predictions_max, 
  length.out = 9
  )

#color palette
color_palette <- grDevices::colorRampPalette(
  c("#2166AC", "white", "#B2182B"))(length(color_breaks))

#differences map
mapview(
  difference_predictions, 
  col.regions = color_palette, 
  at = color_breaks, 
  layer.name = "Suitability difference",
  na.color = "transparent"
  )
```

The divergent palette shows three distinct zones:

- **Blue (negative values)**: indicating areas where the habitat is more suitable for the plant than the butterfly.
- **White (values near zero)**: areas where both species show similar predicted suitability.
- **Red (positive values)**: areas where butterfly suitability exceeds host plant suitability, which may reflect the butterfly's mobility, or just be a model extrapolation artefact.


## Conclusion

The `interaction` dataset illustrates a two-stage species distribution modelling workflow in which the output of one model becomes a predictor for another. By first modelling the host plant *Androsace vitaliana* from abiotic predictors and then using those predictions as a biotic input for the butterfly *Agriades zullichi*, we showed how obligate ecological dependencies can be encoded in a sequential modelling framework.

The resulting suitability maps and difference raster make `interaction` well suited for tutorials on biotic interactions, predictor engineering, spatial cross-validation, and divergent palette visualization in R.

## References

Barea-Azcón, J.M., Benito, B.M., Olivares, F.J., Ruiz, H., Martín, J., García, A.L., & López, R. (2014). Distribution and conservation of the relict interaction between the butterfly *Agriades zullichi* and its larval foodplant (*Androsace vitaliana nevadensis*). *Biodiversity and Conservation*, 23(4), 927–944. https://doi.org/10.1007/s10531-014-0643-4

Benito, B., Lorite, J., & Peñas, J. (2011). Simulating potential effects of climatic warming on altitudinal patterns of key species in Mediterranean-alpine ecosystems. *Climatic Change*, 108, 471–483. https://doi.org/10.1007/s10584-010-0015-3

```{r, echo = FALSE}
unlink("sierra_nevada_env.tif")

```

