---
title: "linaria"
output:
  rmarkdown::html_document:
    toc: true
    toc_title: "Contents"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval      = TRUE,
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.align = "center",
  dpi = 150,
  out.width = "100%"
)
```

## Source Code

You can download the Rmarkdown notebook used to render this article [here](https://github.com/BlasBenito/spatialData/blob/main/vignettes/articles/linaria.Rmd). To download the file, use the button *Download raw file* on the upper-right hand of the `Code` panel.

## Overview

The `linaria` dataset is an `sf` data frame containing presence records of [*Linaria nigricans*](https://powo.science.kew.org/taxon/urn:lsid:ipni.org:names:804793-1) and greenhouses in eastern Andalusia (Spain). The plant's habitat is threatened by the expansion of greenhouses in the region. The dataset also includes 21 environmental predictors and a companion raster data at 400m resolution. `linaria` is well suited for binary classification, multi-response SDMs, and habitat suitability analysis.

## Setup

```{r, include = FALSE}
if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak")
}
library(pak)
pak::pkg_install(
  c(
    "dplyr",
    "sf",
    "stars",
    "mapview",
    "terra",
    "maxnet",
    "spatialData"
  ),
  ask = FALSE
)
```

```{r, message = FALSE, warning = FALSE, echo = FALSE}
library(dplyr)
library(sf)
library(mapview)
library(terra)
library(maxnet)
library(spatialData)

data(
  linaria,
  linaria_responses,
  linaria_predictors,
  package = "spatialData"
)

linaria_color <- "#BF3EFF"
linaria_colors <- grDevices::colorRampPalette(
  colors = c(
    "white", 
    linaria_color,
    "black"
    )
)

greenhouses_color <- "#97FFFF"
greenhouses_colors <- grDevices::colorRampPalette(
  colors = c(
    "white", 
    greenhouses_color,
    "black"
    )
)

env_colors <- colorRampPalette(
  colors = c(
    "white",
    greenhouses_color,
    linaria_color,
    "black"
  )
)
```

## Data Structure

The dataset is an `sf` data frame with `r nrow(linaria)` rows and `r ncol(linaria)` columns.

```{r}
linaria |> 
  head() |> 
  dplyr::glimpse()
```

It has two binary response variables (see `linaria_responses`):

```{r}
linaria_responses
```

Both response variables encode binary outcomes: 1 for confirmed presence, 0 for all other record types.

The map below shows all three record types by colour.

```{r map, message = FALSE}
mapview::mapview(
  linaria |> 
    dplyr::filter(
      linaria_nigricans == 1
    ),
  layer.name = "Linaria nigricans",
  col.regions = linaria_color
) + 
  mapview::mapview(
  linaria |> 
    dplyr::filter(
      greenhouses == 1
    ),
  layer.name = "Greenhouses",
  col.regions = greenhouses_color
)
  
```

The dataset includes `r length(linaria_predictors)` remote sensing, topographic, and climate predictors.

```{r}
linaria_predictors
```
These variables are available as a raster file at 400 m resolution (EPSG:25830), which is loaded with `linaria_extra()`. 

```{r load_raster, message = FALSE, fig.width=6, fig.height=8}
linaria_env <- spatialData::linaria_extra()

plot(
  x = linaria_env, 
  nc = 3,
  col = env_colors(n = 100)
  )
```


## Example Usage

Greenhouse agriculture expands over the habitat of *Linaria nigricans* in eastern Andalusia. Here we assess potential impacts of such expansion on the plant populations by training two separate habitat suitability models using [`maxnet`](https://cran.r-project.org/package=maxnet), one for *L. nigricans* and one for greenhouses, predicting them across the study region, and combining them into a relative risk surface.

### Model Training and Prediction

The code below applies the R version of the popular [MaxEnt](https://biodiversityinformatics.amnh.org/open_source/maxent/) software to model the distribution of *Linaria nigricans* and greenhouses against background data representing the full environmental backdrop of the study region.

```{r fit_sdms}
linaria_training <- sf::st_drop_geometry(linaria)

linaria_model <- maxnet::maxnet(
  p = linaria_training[["linaria_nigricans"]],
  data = linaria_training[, linaria_predictors],
  addsamplestobackground = TRUE #default
)

greenhouses_model <- maxnet::maxnet(
  p = linaria_training[["greenhouses"]],
  data = linaria_training[, linaria_predictors]
)
```


We project each model onto the environmental raster using `terra::predict()` with `type = "cloglog"`, which returns values bounded between 0 and 1 that can be interpreted as occurrence probability.

```{r predict_sdms, message = FALSE}
linaria_prediction <- terra::predict(
  object = linaria_env,
  model  = linaria_model,
  type   = "cloglog",
  na.rm  = TRUE
)
names(linaria_prediction) <- "linaria"

greenhouses_prediction <- terra::predict(
  object = linaria_env,
  model  = greenhouses_model,
  type   = "cloglog",
  na.rm  = TRUE
)
names(greenhouses_prediction) <- "greenhouses"
```

```{r}
mapview::mapview(
  x = linaria_prediction,
  col.regions  = linaria_colors(n = 5),
  layer.name = "Linaria suitability",
  na.color = "transparent"
) +
  mapview(
    linaria |> 
      dplyr::filter(
        linaria_nigricans == 1
      ),
    color = "gray60",
    layer.name = "Linaria presence",
    alpha.regions = 0
  )
```

```{r}
mapview::mapview(
  x = greenhouses_prediction,
  col.regions  = greenhouses_colors(n = 5),
  layer.name = "Greenhouses suitability",
  na.color = "transparent"
) +
  mapview(
    linaria |> 
      dplyr::filter(
        greenhouses == 1
      ),
    color = "gray60",
    layer.name = "Greenhouses presence",
    alpha.regions = 0
  )
```

### Habitat-Loss Risk

We can use these models to obtain a proxy of habitat-destruction risk for the populations *Linaria nigricans* we know about (presence records) and the ones we don't (high suitability regions in `linaria_prediction`).

Let's start with the ones we know about. 

Below we extract the greenhouse suitability scores for all presences of *Linaria nigricans*.

```{r}
risk_sf <- terra::extract(
  x = greenhouses_prediction,
  y = linaria |> 
    dplyr::filter(
      linaria_nigricans == 1
    ) |> 
    dplyr::select(
      geometry
      ),
  bind = TRUE
) |> 
  sf::st_as_sf() 

dplyr::glimpse(risk_sf)
```

The map below uses the output of `terra::extract()` to highlight the known populations at a higher risk by color and size.

```{r, warning = FALSE}
mapview::mapview(
  risk_sf,
  zcol = "greenhouses",
  layer.name = "Risk",
  cex = "greenhouses",
  col.regions = greenhouses_colors(n = 5),
  map.types = c("Esri.WorldImagery")
)
```

Take in mind that the training data represents the 2010's, while the background satellite image is much more recent.

We need to follow several steps to assess potential risks at unknown populations:

- Identify the cell IDs of the known populations.
- Convert the linaria and greenhouse predictions to a data frame with cell IDs.
- Remove the cell IDs of the known populations.



```{r}
#create buffer of 500m around presences
linaria_presence_buffer <- terra::buffer(
  x = linaria |> 
    dplyr::filter(
      linaria_nigricans == 1
    ) |> 
    terra::vect(),
  width = 500
)

#extract cell IDs inside the buffer
linaria_presence_cells <- terra::extract(
  x = linaria_prediction,
  y = linaria_presence_buffer,
  cells = TRUE,
  ID = FALSE,
  na.rm = TRUE
) |> 
  na.omit()

dplyr::glimpse(linaria_presence_cells)
```

Below we create a data frame with cell IDs and their coordinates, and predicted suitabilities for linaria and greenhouses.

```{r}
potential_risk <- dplyr::bind_cols(
  as.data.frame(
    x = linaria_prediction,
    cells = TRUE,
    xy = TRUE
  ),
  as.data.frame(
    x = greenhouses_prediction
  )
)

dplyr::glimpse(potential_risk)
```

We can now remove the cells belonging to known populations, and filter the data frame to those cases with high sutiability for linaria and greenhouses.

```{r}
potential_risk <- potential_risk |> 
  #remove cells with known populations
  dplyr::filter(
    !cell %in% linaria_presence_cells$cell
  ) |> 
  #remove cases with low linaria and greenhouse suitability
  dplyr::filter(
    linaria > 0.5,
    greenhouses > stats::median(greenhouses)
  ) |> 
  #to sf
  sf::st_as_sf(
    coords = c("x", "y"),
    crs = sf::st_crs(linaria)
  )
```

The resulting map highlights sites with high-suitability for *Linaria nigricans* that might be at risk due to greenhouse expansion.

```{r, warning = FALSE}
mapview::mapview(
  potential_risk,
  zcol = "greenhouses",
  layer.name = "Potential risk",
  cex = "greenhouses",
  col.regions = greenhouses_colors(n = 5),
  map.types = c("Esri.WorldImagery")
) + 
  mapview::mapview(
    linaria |> 
      dplyr::filter(
        linaria_nigricans == 1
      ),
    layer.name = "Known populations",
    alpha.regions = 0,
    col.regions = "white",
    col = "white",
    cex = 0.5
  )
```


## Conclusion

The `linaria` dataset provides a rich testbed for multi-response binary classification: one response captures species habitat suitability, the other captures the spatial footprint of the main threatening land use. This structure is ideal for comparative SDM workflows, land-use conflict analyses, or any task requiring simultaneous modelling of species and threat distributions.

## References

Benito, B.M., Martínez-Ortega, M.M., Munoz, L.M., Lorite, J. & Penas, J. (2009). Assessing extinction-risk of endangered plants using species distribution models: a case study of habitat depletion caused by the spread of greenhouses. *Biodiversity and Conservation*, 18(9), 2509–2520. https://doi.org/10.1007/s10531-009-9604-8

Peñas, J., Benito, B., Lorite, J., et al. (2011). Habitat fragmentation in arid zones: a case study of *Linaria nigricans* under land use changes (SE Spain). *Environmental Management*, 48, 168–176. https://doi.org/10.1007/s00267-011-9663-y

Benito, B.P.d., Peñas, J.G.d. (2008). Greenhouses, land use change, and predictive models: MaxEnt and Geomod working together. In: Paegelow, M., Olmedo, M.T.C. (eds) Modelling Environmental Dynamics. Environmental Science and Engineering. Springer, Berlin, Heidelberg. https://doi.org/10.1007/978-3-540-68498-5_11

