---
title: "linaria"
output:
  rmarkdown::html_document:
    toc: true
    toc_title: "Contents"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

## Overview

The `linaria` dataset, built from *Benito et al. (2009)*, contains three types of records:

- **Species presences** (*Linaria nigricans* occurrence sites)
- **Greenhouse presences** (competing land-use sites)
- **Background points** (random samples from the study extent)

Combined with 21 environmental predictors (Landsat reflectance, climate, topography) at 400 m resolution, the dataset is well suited for binary classification, multi-response SDMs, and habitat suitability analysis.

## Setup

```{r, include = FALSE}
if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak")
}
pak::pkg_install(
  c(
    "sf",
    "stars",
    "mapview",
    "dplyr",
    "terra",
    "ggplot2",
    "patchwork",
    "viridis",
    "maxnet"
  ),
  ask = FALSE
)
```

```{r, message = FALSE, warning = FALSE, echo = FALSE}
library(dplyr)
library(terra)
library(sf)
library(mapview)
library(ggplot2)
library(patchwork)
library(viridis)
library(maxnet)
library(spatialData)

data(
  linaria,
  linaria_responses,
  linaria_predictors,
  package = "spatialData"
)
```

## Data Structure

The dataset is an `sf` data frame with `r nrow(linaria)` rows and `r ncol(linaria)` columns.

```{r}
linaria |> 
  head() |> 
  dplyr::glimpse()
```

It has two binary response variables (see `linaria_responses`):

```{r}
linaria_responses
```

Both response variables encode binary outcomes: 1 for confirmed presence, 0 for all other record types.

The map below shows all three record types by colour.

```{r map, message = FALSE}
linaria_mapped <- linaria |>
  dplyr::mutate(
    record_type = dplyr::case_when(
      linaria_nigricans == 1L ~ "Linaria nigricans",
      greenhouses == 1L      ~ "Greenhouse",
      TRUE                   ~ "Background"
    )
  )

mapview::mapview(
  linaria_mapped,
  zcol = "record_type",
  layer.name = "Record type",
  col.regions = c("gray60", "darkgreen", "violet")
)
```

The dataset includes `r length(linaria_predictors)` remote sensing, topographic, and climate predictors.

```{r}
linaria_predictors
```

## Example Usage

Greenhouse agriculture and *Linaria nigricans* compete for the same lowland territory in eastern Andalusia. Following an approach similar to *Benito et al. (2009)*, we fit two separate habitat suitability models using [`maxnet`](https://cran.r-project.org/package=maxnet), one for *L. nigricans* and one for greenhouses, predict them across the study region, and combine them into an extinction-risk surface by multiplying the two suitability layers and rescaling to 0–100.

### Environmental raster

The environmental raster (same extent, 400 m resolution, EPSG:25830) is loaded with `linaria_extra()`. When the vignette renders, the raster file `linaria_env.tif` is read from disk without re-downloading.

```{r load_raster, message = FALSE}
linaria_env <- spatialData::linaria_extra()
```

```{r}
plot(linaria_env)
```

### Fitting SDMs with `maxnet`

Each `maxnet()` call treats rows where the response equals 1 as presences and all other rows (including the opposite response's presences and background points) as background. This pseudo-absence design lets us model each response against the full environmental backdrop of the study region.

```{r fit_sdms}
training_df <- sf::st_drop_geometry(linaria)

linaria_model <- maxnet::maxnet(
  p = training_df[["linaria_nigricans"]],
  data = training_df[, linaria_predictors]
)

greenhouses_model <- maxnet::maxnet(
  p = training_df[["greenhouses"]],
  data = training_df[, linaria_predictors]
)
```


We project each model onto the environmental raster using `terra::predict()` with `type = "cloglog"`, which returns values bounded between 0 and 1 that can be interpreted as occurrence probability. The option `clamp = TRUE` restricts predictions to the environmental envelope seen during training.

```{r predict_sdms, message = FALSE}
linaria_prediction <- terra::predict(
  object = linaria_env,
  model  = linaria_model,
  type   = "cloglog",
  clamp  = TRUE,
  na.rm  = TRUE
)
names(linaria_prediction) <- "linaria"

greenhouses_prediction <- terra::predict(
  object = linaria_env,
  model  = greenhouses_model,
  type   = "cloglog",
  clamp  = TRUE,
  na.rm  = TRUE
)
names(greenhouses_prediction) <- "greenhouses"
```

```{r plot_suitability}
plot(
  linaria_prediction,
  col  = viridis::turbo(100),
  main = "Linaria nigricans suitability"
)

plot(
  greenhouses_prediction,
  col  = viridis::turbo(100),
  main = "Greenhouse suitability"
)
```

### Extinction-risk composite

The relative extinction-risk surface is the product of the two suitability layers. Areas with high scores represent places where *L. nigricans* is likely to occur *and* where greenhouses are likely to expand, indicating high-risk zones.

```{r linaria_risk}
linaria_risk <- terra::stretch(
  x = linaria_prediction * greenhouses_prediction, 
  minv = 0, 
  maxv = 100
  )
```

```{r plot_risk, fig.height = 5}
mapview(
  linaria_risk,
  col.regions  = viridis::turbo(100),
  layer.name = "Relative risk (0–100)"
) + 
  mapview(
    linaria |> 
      dplyr::filter(
        linaria_nigricans == 1
      ),
    layer.name = "Linaria presence",
    alpha.regions = 0
  )
```
The original data is from 2008. Notice how several presence records in the high-risk areas are already covered by greenhouses.


## Conclusion

The `linaria` dataset provides a compact but ecologically rich testbed for multi-response binary classification: one response captures species habitat suitability, the other captures the spatial footprint of the main threatening land use. This structure is ideal for comparative SDM workflows, land-use conflict analyses, or any task requiring simultaneous modelling of species and threat distributions.

The extinction-risk analysis above shows how combining two MaxEnt predictions into a spatially explicit composite can pinpoint the areas where conservation action is most urgent — the approach introduced by Benito et al. (2009) for *Linaria nigricans* in Andalusia.

```{r, echo = FALSE}
unlink("linaria_env.tif")

```

