---
title: "neanderthal"
output:
  rmarkdown::html_document:
    toc: true
    toc_title: "Contents"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval      = TRUE,
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.align = "center",
  dpi = 150,
  out.width = "100%"
)
```

## Source Code

You can download the Rmarkdown notebook used to render this article [here](https://github.com/BlasBenito/spatialData/blob/main/vignettes/articles/neanderthal.Rmd). To download the file, use the button *Download raw file* on the upper-right hand of the `Code` panel.

## Overview

The `neanderthal` dataset contains records of Neanderthal archaeological sites from Marine Isotope Stage 5e (Last Interglacial, ~130,000-115,000 years ago), combined with 25 paleoclimate and topographic predictor variables. This dataset was built to support binary classification, species distribution modelling, and habitat suitability analysis.

## Setup

The code below loads the R libraries and example data required to run this tutorial.

```{r, include = FALSE}
if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak")
}
pak::pkg_install(
  c(
    "dplyr",
    "sf",
    "mapview",
    "terra",
    "viridis",
    "mgcv",
    "ggplot2",
    "patchwork",
    "collinear",
    "spatialData"
  ),
  ask = FALSE
)
```

```{r, message = FALSE, warning = FALSE, echo = FALSE}
library(dplyr)
library(sf)
library(mapview)
library(terra)
library(mgcv)
library(ggplot2)
library(patchwork)
library(viridis)
library(collinear)
library(spatialData)

data(
  neanderthal,
  neanderthal_response,
  neanderthal_predictors,
  package = "spatialData"
)

neanderthal_colors <- grDevices::hcl.colors(
  palette = "Zissou 1",
  n = 10
)
neanderthal_color <- neanderthal_colors[5]

env_colors <- grDevices::hcl.colors(
  palette = "YlGnBu",
  n = 10
)
```



## Data Structure

The dataset is an `sf` data frame with `r nrow(neanderthal)` rows and `r ncol(neanderthal)` columns, and no missing data.

```{r}
dplyr::glimpse(neanderthal)
```

The response variable is `presence` (see `neanderthal_response`), a integer binary variable indicating Neanderthal presence (1) or pseudo-absence (0).

```{r}
table(neanderthal$presence)
```
The locations of the presence and background records are mapped below. 

```{r map, message = FALSE}
mapview::mapview(
  neanderthal |> 
    dplyr::mutate(
      presence = as.factor(presence)
    ),
  zcol = "presence",
  layer.name = "Presence",
  col.regions = c("gray50", neanderthal_color)
)
```

The dataset includes 25 palaeoclimatic (last interglacial) and topographic predictors.

```{r}
neanderthal_predictors
```
There is a companion raster dataset loaded via `neanderthal_extra()`.

```{r load_raster, message = FALSE, fig.width=6, fig.height=8}
neanderthal_env <- spatialData::neanderthal_extra()

plot(
  x = neanderthal_env, 
  nc = 3,
  col = env_colors,
  )
```

## Example Usage

In this section we use `neanderthal` to model habitat suitability using an ensemble of univariate GAM models.

Each univariate GAM model in our ensemble is going to look like this:

```{r, eval = FALSE}
m <- mgcv::gam(
  formula = presence ~ s(bio1, k = ceiling(sum(neanderthal$presence)/10)),
  data = sf::st_drop_geometry(neanderthal),
  family = stats::quasibinomial(link = "logit"),
  weights = collinear::case_weights(neanderthal$presence),
  REML = TRUE,
  select = TRUE
)
```

- `mgcv::gam()`: GAM models replace the linear terms used in GLMs with smoothing splines (`s()`) that better represent non-linear relationships between presence and environment.

- `k = ceiling(sum(neanderthal$presence)/10)`: The basis dimension ensures that each smooth well within the budget allowed by the number of presences. 

- `family = stats::quasibinomial(link = "logit")`: The quasibinomial family accounts for the overdispersion introduced by pseudo-absences and case weights.

- `REML = TRUE` and `select = TRUE`: the actual complexity is penalized further, and uninformative terms can be shrunk to zero.

- `weights = collinear::case_weights(neanderthal$presence)`: inverse-frequency weights so that both classes, no matter how imbalanced, contribute equally to the model fit.

```{r}
sample_weights <- collinear::case_weights(x = neanderthal$presence)
table(sample_weights)
```


**Weighted Ensemble**: The `neanderthal` data has `r sum(neanderthal$presence)` presences. Fitting a model with all predictors at once might lead to overparameterization and overfitting. To avoid this potential issue, we will fit univariate GAM models between presence and each environmental predictor, and then **ensemble** their predictions via weighted average, using the models' AUC scores as weights.


### Multicollinearity Filtering

There are many redundant predictors in `neanderthal`. To mitigate this issue, here we apply [`collinear::collinear()`](https://blasbenito.github.io/collinear/reference/collinear.html) to select a non-redundant subset of the most relevant predictors. The function ranks all predictors by the AUC of univariate binomial GAM models with case weights (very similar to the ones we'll use for the habitat suitability model!), and then applies a multicollinearity filtering that preserves the most important ones.

```{r}
neanderthal_selection <- collinear::collinear(
  df = neanderthal,
  responses = neanderthal_response,
  predictors = neanderthal_predictors,
  f = collinear::f_binomial_gam
)
```

The names of the selected predictors are in the vector below.

```{r}
neanderthal_selection$presence$selection
```

### Model Fitting

The code below fits the ensemble of univariate GAM models. 

```{r}
models <- lapply(
  X = neanderthal_selection$presence$selection,
  FUN = function(x){
    
    mgcv::gam(
      formula = as.formula(
        paste0(
          "presence ~ s(", 
          x, ", 
          k = ", 
          floor(sum(neanderthal$presence)/10), 
          ")"
          )
        ),
      data = sf::st_drop_geometry(neanderthal),
      family = stats::quasibinomial(link = "logit"),
      weights = sample_weights,
      REML = TRUE,
      select = TRUE
    )
    
  }
)

names(models) <- neanderthal_selection$presence$selection
```

The following plots show the predicted probability of Neanderthal presence for each univariate GAM model. Rug marks along the x-axis indicate observed data values.

```{r, fig.width=7, fig.height=8}
response_plots <- lapply(names(models), function(var) {

  #extract individual univariate model
  model <- models[[var]]
  df <- sf::st_drop_geometry(neanderthal)

  #prediction grid
  newdata <- data.frame(
    x = seq(
      from = min(df[[var]]), 
      to = max(df[[var]]), 
      length.out = 200)
    )
  names(newdata) <- var

  #predict on link scale with standard errors
  pred <- predict(
    object = model,
    newdata = newdata, 
    type = "link", 
    se.fit = TRUE
    )

  #back-transform to probability scale with 95% CI
  newdata$fit <- stats::plogis(pred$fit)
  newdata$lower <- stats::plogis(pred$fit - 1.96 * pred$se.fit)
  newdata$upper <- stats::plogis(pred$fit + 1.96 * pred$se.fit)

  #plot response curve
  ggplot(newdata, aes(x = .data[[var]])) +
    geom_ribbon(
      aes(ymin = lower, ymax = upper),
      fill = "gray50",
      alpha = 0.25
    ) +
    geom_line(
      aes(y = fit),
      color = "red4",
      linewidth = 0.8
    ) +
    geom_rug(
      data = df[df$presence == 1, ],
      aes(x = .data[[var]]),
      sides = "t",
      color = "red4",
      alpha = 0.3,
      length = unit(0.1, "npc")
    ) +
    geom_rug(
      data = df[df$presence == 0, ],
      aes(x = .data[[var]]),
      sides = "b",
      color = "gray50",
      alpha = 0.3,
      length = unit(0.05, "npc")
    ) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(x = var, y = "P(presence)") +
    theme_minimal(base_size = 11)
})

patchwork::wrap_plots(response_plots, ncol = 2)
```

### Model Evaluation

In this section we compute the partial AUC scores of each univariate model, and the global AUC of the ensemble. A more honest cross-validation evaluation would work better here, but it is left out for simplicity.

The partial AUC scores will be used as weights when building the ensemble. Here these scores are computed via [collinear::score_auc()](https://blasbenito.github.io/collinear/reference/score_auc.html), but any other function able to compute AUC will do here.

```{r}
model_weights <- lapply(
  X = models,
  FUN = function(x){
    
    collinear::score_auc(
      o = neanderthal[["presence"]],
      p = stats::predict(x, type = "response")
    )
    
  }
) |> 
  unlist()

names(model_weights) <- neanderthal_selection$presence$selection
```

```{r, echo = FALSE}
round(as.data.frame(model_weights), digits = 4)
```

To compute the intrinsic AUC of the model, we first generate the predictions of the univariate GAM models, and then average these predictions weighted by their partial AUC scores, and finally compare them with the observations.


```{r}
#extract predictions from each model
predictions <- sapply(
  X = models,
  FUN = function(x) {
    stats::predict(x, type = "response")
  }
)

#compute weighted average predictions
weighted_predictions <- apply(
  X = predictions,
  MARGIN = 1,
  FUN = stats::weighted.mean,
  w = model_weights
)

#compute ensemble AUC
auc <- collinear::score_auc(
  o = neanderthal[["presence"]],
  p = weighted_predictions
)

auc
```

The global AUC score indicates the probability of any presence to have a higher habitat suitability than any pseudoabsence.

With a validated ensemble in hand, we can now map Neanderthal habitat across the region.

### Habitat Suitability Map


The code below predicts the univariate GAMs over `neanderthal_env`, and then computes their weighted mean to produce the habitat suitability map.

```{r, fig.width=8, fig.height=4}
#predict each model on a separate raster object
univariate_predictions <- lapply(
  X = names(models),
  FUN = function(x) {
    terra::predict(
      object = neanderthal_env,
      model = models[[x]],
      type = "response",
      na.rm = TRUE
    )
  }
) |> 
  terra::rast()

#compute weighted mean of individual predictions
ensemble_average <- terra::weighted.mean(
  x = univariate_predictions,
  w = model_weights
)

#weighted standard deviation of the ensemble
ensemble_deviation <- terra::app(
  x = univariate_predictions,
  fun = function(x) {
    sqrt(sum(model_weights * (x - weighted.mean(x, model_weights))^2) / sum(model_weights))
    }
  )

#plot suitability map
par(mfrow = c(1, 2))
plot(
  x = ensemble_average,
  col = viridis::turbo(n = 100),
  main = "Weighted Average"
)
plot(
  x = ensemble_deviation,
  main = "Weighted Standard Deviation"
)
```

Finally, we combine both maps into a single visualization using a **whitening** approach: the suitability color scale is progressively bleached toward white as uncertainty increases. Vivid colors indicate confident predictions, while pale or white areas flag regions where models disagree.

```{r, fig.width=7, fig.height=5}

#combine into a data frame
bi_df <- terra::as.data.frame(
  x = c(ensemble_average, ensemble_deviation),
  xy = TRUE
)
names(bi_df)[3:4] <- c("suitability", "uncertainty")

#normalize both variables to [0, 1]
bi_df$suit_norm <- (bi_df$suitability - min(bi_df$suitability)) /
  (max(bi_df$suitability) - min(bi_df$suitability))
bi_df$unc_norm <- (bi_df$uncertainty - min(bi_df$uncertainty)) /
  (max(bi_df$uncertainty) - min(bi_df$uncertainty))

#map suitability to turbo colors
turbo_palette <- viridis::turbo(100)
base_colors <- col2rgb(
  turbo_palette[ceiling(bi_df$suit_norm * 99) + 1]
) / 255

#blend toward white based on uncertainty
r <- base_colors["red", ] * (1 - bi_df$unc_norm) + bi_df$unc_norm
g <- base_colors["green", ] * (1 - bi_df$unc_norm) + bi_df$unc_norm
b <- base_colors["blue", ] * (1 - bi_df$unc_norm) + bi_df$unc_norm
bi_df$fill_color <- rgb(r, g, b)

#build map
bi_map <- ggplot2::ggplot(
  data = bi_df,
  mapping = ggplot2::aes(x = x, y = y, fill = fill_color)
) +
  ggplot2::geom_raster() +
  ggplot2::scale_fill_identity() +
  ggplot2::coord_quickmap() +
  ggplot2::theme_minimal() +
  ggplot2::labs(x = "", y = "")

#build whitening legend
n <- 100
legend_df <- expand.grid(
  suitability = seq(0, 1, length.out = n),
  uncertainty = seq(0, 1, length.out = n)
)
leg_colors <- col2rgb(
  turbo_palette[ceiling(legend_df$suitability * 99) + 1]
) / 255
legend_df$fill <- rgb(
  leg_colors["red", ] * (1 - legend_df$uncertainty) + legend_df$uncertainty,
  leg_colors["green", ] * (1 - legend_df$uncertainty) + legend_df$uncertainty,
  leg_colors["blue", ] * (1 - legend_df$uncertainty) + legend_df$uncertainty
)

bi_legend <- ggplot2::ggplot(
  data = legend_df,
  mapping = ggplot2::aes(
    x = suitability,
    y = uncertainty,
    fill = fill
  )
) +
  ggplot2::geom_raster() +
  ggplot2::scale_fill_identity() +
  ggplot2::scale_x_continuous(expand = c(0, 0)) +
  ggplot2::scale_y_continuous(expand = c(0, 0)) +
  ggplot2::labs(
    x = "Suitability \u2192",
    y = "Uncertainty \u2192"
  ) +
  ggplot2::theme_minimal(base_size = 8) +
  ggplot2::theme(
    axis.text = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank(),
    panel.grid = ggplot2::element_blank(),
    panel.border = ggplot2::element_rect(fill = NA, color = "black", linewidth = 0.5)
  )

#compose map and legend
bi_map + patchwork::inset_element(
  p = bi_legend,
  left = 0.70,
  bottom = -0.1,
  right = 1,
  top = 0.25
)

```

The map suggests that during the Last Interglacial, the most suitable Neanderthal habitat was concentrated in southern Europe and parts of the Near East, with model confidence decreasing toward range margins and data-sparse regions.

## Conclusion

The ensemble of univariate GAMs, weighted by cross-validated AUC, provides a robust picture of Neanderthal habitat suitability during the Last Interglacial. Whitening the ensemble predictions with their standard deviation surfaces the areas of genuine uncertainty. SDM tools developed for modern species translate naturally to palaeontological data — and `neanderthal` is a good place to experiment with that.

```{r, echo = FALSE}
unlink("neanderthal_env.tif")

```

