---
title: "Neanderthal: Presence-Absence Data for Species Distribution Modeling"
output:
  rmarkdown::html_document:
    toc: true
    toc_title: "Contents"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

## Overview

The `neanderthal` dataset contains **245 records** of Neanderthal archaeological sites from Marine Isotope Stage 5e (the Last Interglacial, ~130,000-115,000 years ago) in Europe and the Near East. This dataset is designed for presence-absence species distribution modeling, combining presence records with pseudo-absence points.

**Data source:** Benito, B. M., et al. (2017). The ecological niche and distribution of Neanderthals during the Last Interglacial. *Journal of Biogeography*, 44(1), 51-61. [doi:10.1111/jbi.12845](https://doi.org/10.1111/jbi.12845)

## Loading the Data

```{r setup}
library(spatialData)

# Load the main dataset
data(neanderthal)

# Load helper vectors
data(neanderthal_response)
data(neanderthal_predictors)

# View helper vectors
neanderthal_response
neanderthal_predictors
```

## Data Structure

The `neanderthal` dataset is an **sf data frame** with point geometries in WGS84 (EPSG:4326):

```{r structure}
class(neanderthal)
dim(neanderthal)
str(neanderthal, give.attr = FALSE)
```

## Response Variable

The response variable is `presence`, a binary indicator:

- **1** = Archaeological presence record (confirmed Neanderthal site)
- **0** = Pseudo-absence point (environmentally similar location without evidence)

```{r response}
table(neanderthal$presence)
```

## Predictor Variables

The dataset includes **25 predictor variables** organized into two categories:

### Bioclimatic Variables (19 variables)

Standard WorldClim bioclimatic variables representing paleoclimate reconstructions for MIS5e:

| Variable | Description |
|----------|-------------|
| `bio1` | Annual mean temperature |
| `bio2` | Mean diurnal range |
| `bio3` | Isothermality |
| `bio4` | Temperature seasonality |
| `bio5` | Max temperature of warmest month |
| `bio6` | Min temperature of coldest month |
| `bio7` | Temperature annual range |
| `bio8` | Mean temperature of wettest quarter |
| `bio9` | Mean temperature of driest quarter |
| `bio10` | Mean temperature of warmest quarter |
| `bio11` | Mean temperature of coldest quarter |
| `bio12` | Annual precipitation |
| `bio13` | Precipitation of wettest month |
| `bio14` | Precipitation of driest month |
| `bio15` | Precipitation seasonality |
| `bio16` | Precipitation of wettest quarter |
| `bio17` | Precipitation of driest quarter |
| `bio18` | Precipitation of warmest quarter |
| `bio19` | Precipitation of coldest quarter |

### Topographic Variables (6 variables)

| Variable | Description |
|----------|-------------|
| `topo_aspect` | Terrain aspect |
| `topo_diversity_local` | Local topographic diversity |
| `topo_diversity` | Regional topographic diversity |
| `topo_elev` | Elevation |
| `topo_slope` | Terrain slope |
| `topo_wetness` | Topographic wetness index |

### Summary Statistics

```{r predictors}
summary(neanderthal[, neanderthal_predictors])
```

## Interactive Map

The map below shows the spatial distribution of presence (blue) and pseudo-absence (red) records:
```{r map, message=FALSE}
library(mapview)

mapview(
  neanderthal,
  zcol = "presence",
  layer.name = "Presence"
)
```

## Citation

If you use this dataset, please cite:

> Benito, B. M., et al. (2017). The ecological niche and distribution of Neanderthals during the Last Interglacial. *Journal of Biogeography*, 44(1), 51-61. doi:10.1111/jbi.12845
