---
title: "Quercus: European Oak Species Distribution Data"
output:
  rmarkdown::html_document:
    toc: true
    toc_title: "Contents"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

## Overview

The `quercus` dataset contains presence and absence records of eight European Quercus (oak) species combined with 31 environmental predictor variables covering bioclimatic, topographic, vegetation, and human impact dimensions.

Its main features are:

- *Complete case study* - No missing values, clean data ready for analysis.
- *Multi-class classification* - Eight oak species plus absence points (9 levels).
- *European scope* - Records spanning western, central, and southern Europe.
- *Ecological predictors* - Climate, topography, vegetation indices, land cover, and human footprint.

This dataset was built to support a range of analytical approaches, such as binary and multi-class classification, niche modelling, and niche overlap analysis.

## Setup

The following R libraries are required to run this tutorial:

```{r, include = FALSE}
packages <- c("sf", "mapview", "DT", "dplyr", "collinear", "rpart", "rpart.plot", "viridis", "terra")
installed <- vapply(packages, requireNamespace, logical(1), quietly = TRUE)
if (any(!installed)) install.packages(packages[!installed])
```

```{r, message = FALSE, warning = FALSE, echo = FALSE}
library(spatialData)
library(sf)
library(terra)
library(mapview)
library(DT)
library(dplyr)
library(collinear)
library(rpart)
library(rpart.plot)
library(viridis)
```

The code below loads the example data.

```{r setup}
data(
  quercus,
  quercus_response,
  quercus_predictors,
  package = "spatialData"
)
```

## Data Structure

The dataset is an `sf` data frame with `r nrow(quercus)` rows and `r ncol(quercus)` columns, and no missing data. The first 10 records and all columns but `geometry` are shown below.

```{r}
DT::datatable(
  data = head(sf::st_drop_geometry(quercus), n = 10),
  rownames = FALSE
)
```

## Response Variable

The response variable is `species`, a categorical variable with 9 levels (8 species + absence):

```{r}
table(quercus$species)
```

## Predictor Variables

The dataset includes **31 predictor variables** organized into six categories:

| Category | Variables | Source |
|----------|-----------|--------|
| Bioclimatic (17) | `bio1`, `bio2`, ..., `bio19` (excluding bio8, bio9) | WorldClim |
| NDVI (4) | `ndvi_average`, `ndvi_maximum`, `ndvi_minimum`, `ndvi_range` | MODIS |
| Solar radiation (4) | `sun_rad_average`, `sun_rad_maximum`, `sun_rad_minimum`, `sun_rad_range` | WorldClim |
| Land cover (3) | `landcover_veg_bare`, `landcover_veg_herb`, `landcover_veg_tree` | MODIS MOD44B |
| Topographic (2) | `topo_slope`, `topographic_diversity` | SRTM |
| Human impact (1) | `human_footprint` | Venter et al. 2016 |


## Interactive Map

The map below shows the spatial distribution of oak species and absence points across Europe:

```{r map, message = FALSE}
mapview::mapview(
  quercus,
  zcol = "species",
  layer.name = "Species"
)
```

## Example Usage

This example shows how `quercus` can be used to identify the environmental conditions that distinguish oak species using multicollinearity filtering and a decision tree.

### Multicollinearity Filtering

Many predictors in this dataset are correlated (e.g., `bio5`, `bio6`, `bio7` are all derived from temperature extremes). The `collinear::collinear()` function from the R package [`collinear`](https://blasbenito.github.io/collinear/) selects a non-redundant subset that maximizes predictive power for the response variable.

```{r}
predictors_selection <- collinear::collinear(
  df = quercus,
  responses = "species",
  predictors = quercus_predictors,
  f = collinear::f_categorical_rf,
  quiet = TRUE
)

predictors_selection$species$selection
```

### Decision Tree Model

Species occurrence counts are imbalanced — *Q. robur* has 1,660 records while *Q. suber* has only 58. Without correction, rare species get absorbed into leaves dominated by common ones. The `collinear::case_weights()` function generates inverse-frequency weights so that each species contributes equally to the tree splits.

```{r}
m <- rpart::rpart(
  formula = predictors_selection$species$formulas$classification,
  data = sf::st_drop_geometry(quercus),
  method = "class",
  weights = collinear::case_weights(x = quercus$species)
)
```

### Tree Visualization

The decision tree below shows which environmental variables and thresholds best separate oak species. Each split represents a condition (e.g., "bio6 < -2.3"), and each leaf shows the most likely species at that combination of conditions.

```{r, fig.width = 10, fig.height = 7}
rpart.plot::rpart.plot(
  m,
  type = 0,
  extra = 8,
  fallen.leaves = TRUE,
  roundint = FALSE,
  cex = 0.7,
  box.palette = as.list(viridis::turbo(n = 9, alpha = 0.5)),
  main = "Decision Tree: European Oak Species Classification"
)
```

The figure shows how different combinations of environmental conditions lead to the dominance of each Quercus species.

### Prediction Map

Assigning tree predictions back to the spatial data reveals the locations where each species has the highest probability to appear.


```{r, message = FALSE}
quercus$predicted <- predict(
  object = m,
  newdata = sf::st_drop_geometry(quercus),
  type = "class"
)

mapview::mapview(
  quercus,
  zcol = "predicted",
  layer.name = "Predicted species",
  col.regions = viridis::turbo(
    n = length(unique(quercus$predicted))
  )
)
```

### Species Probability Map

The package provides the raster dataset `quercus_env` to predict models based on the `quercus` data.

```{r}
quercus_env <- spatialData::quercus_extra()
class(quercus_env)
names(quercus_env)
```
```{r, fig.width=5, fig.height=5}
plot(
  x = quercus_env[["bio1"]],
  col = viridis::turbo(n = 100)
  )
```

The map below shows the fitted probability of *Quercus petraea* across Europe.

```{r, fig.width=5, fig.height=5}
quercus_probability <- predict(
  object = quercus_env,
  model = m,
  type = "prob",
  na.rm = TRUE
)

plot(
  x = quercus_probability[["Quercus.petraea"]],
  col = viridis::viridis(n = 5)
  )
```

The decision tree provides interpretable rules for species separation, but more flexible methods like random forests or gradient boosting may achieve higher accuracy and more visually pleasant results. 
