---
title: "Mesoamerican tree richness across the Americas: conditional inference trees for biodiversity patterns"
output:
  rmarkdown::html_document:
    toc: true
    toc_title: "Contents"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

Tree species richness in the Americas is not evenly distributed — Mesoamerican forests are global hotspots of tree diversity, yet that richness declines sharply toward drier or colder regions. Understanding what drives those gradients calls for a dataset with both broad geographic coverage and diverse climate predictors.

## Overview

The `trees` dataset focuses on the distribution of the tree species present in Mesoamerica across North and South America. The presence of Mesoamerican trees was obtained from the **Tree Biodiversity Network (BIOTREE-NET)** dataset (project now defunct). The distribution of these species across the Americas was derived from **GBIF**. The data represents *richness of Mesoamerican trees in the Americas* for 3,373 hexagonal grid cells across the Americas, combined with 50 environmental predictor variables from different sources.

Its main features are:

- *Americas scope*: Hexagonal cells covering longitudes -125.3° to -34.3° and latitudes -34.4° to 49.9°.
- *Single response*: `trees`, an integer count of tree species richness per hexagonal cell.
- *Rich environmental predictors*: 50 predictors spanning 10 categories (climate, soil, vegetation, geography, and more).

We designed it to support regression modelling, multicollinearity filtering, and spatial analysis of tree diversity patterns.

## Setup

```{r, include = FALSE}
if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak")
}
pak::pkg_install(
  c(
    "sf",
    "mapview",
    "dplyr",
    "collinear",
    "partykit",
    "viridis"
  ),
  ask = FALSE
)
```

The following R libraries and example data are required to run this tutorial.

```{r, message = FALSE, warning = FALSE, echo = FALSE}
library(spatialData)
library(sf)
library(mapview)
library(dplyr)
library(collinear)
library(partykit)
library(viridis)

data(
  trees,
  trees_response,
  trees_predictors,
  package = "spatialData"
)
```

## Description

The dataset is an `sf` data frame with `r nrow(trees)` rows and `r ncol(trees)` columns, and 989 cells with `NA` in the `trees` response variable (cells where environmental data is valid but no tree species were found in the relevant databases). The first 10 records and all columns but `geometry` are shown below.

```{r}
trees |>
  head(n = 10) |>
  dplyr::glimpse()
```

The hexagonal grid cells cover the Americas between approximately 50°N and 35°S. The map below shows the grid coloured by tree richness.

```{r, warning = FALSE}
mapview::mapview(
  trees,
  zcol = "trees",
  layer.name = "trees",
  col.regions = viridis::turbo(
    n = 100,
    direction = 1
  ),
  color = NULL,
  at = c(1, 10, 100, 1000, 10000, 20000)
)
```

## Response Variable

The response variable `trees` (named in the vector `trees_response`) is an **integer count** of tree species richness per hexagonal cell, derived from GBIF and BIOTREE-NET occurrence records acquired *circa* 2013.

```{r}
summary(trees$trees)
```


The dataset includes **50 numeric predictors**, named in the vector `trees_predictors`.

```{r}
trees |> 
  sf::st_drop_geometry() |> 
  dplyr::select(
    dplyr::all_of(trees_predictors)
  ) |> 
  head(n = 5) |> 
  dplyr::glimpse()
```

## Raw Data

You can download the raw presence data with `trees_extra()`.

```{r}
trees_presence <- spatialData::trees_extra()
head(trees_presence)
```


## Example Usage

In this example we use `trees` to assess how climate extremes drive tree richness across the Americas.

- *Log Transformation*: to facilitate the visualization of model results.
- *Variable Selection*: multicollinearity analysis of the selected environmental variables.
- *Conditional Inference Trees*: fits a permutation-based tree model to explain tree richness patterns.
- *Prediction Map*: maps the model predictions back to the hexagonal grid.

### Log Transformation

Before proceeding with the variable selection, we log-transform the response variable `trees` into a new column `log_trees` to facilitate the visualization of model results. 

Tree species richness spans several orders of magnitude across the Americas, producing a strongly skewed distribution that makes terminal node summaries in the conditional inference tree difficult to read. While log-transforming count data is no longer recommended for linear models, it has no meaningful effect on split selection in `ctree()`, which relies on rank-based permutation tests that are invariant to monotone transformations of the response.

```{r}
trees <- trees |> 
  dplyr::mutate(
    log_trees = log(trees)
  )
```

### Variable Selection

Now let's identify which extreme-climate predictors are worth keeping.

The predictors indicating climate extreme values are identified with the suffixes `_min` and `_max` in `trees_predictors`, with the exception of rainfall, where `rainfall_driest_month` and `rainfall_wettest_month` represent the monthly minimum and maximum respectively. 

The code below selects these predictors, and excludes `ndvi_*`.


```{r}
trees_extremes <- grep(
  pattern = "^(?!.*ndvi).*(min|max|driest_month|wettest_month)",
  x = colnames(trees),
  perl = TRUE,
  value = TRUE
)

trees_extremes
```

Once the candidate pool of variables is selected, we use [`collinear::collinear()`](https://blasbenito.github.io/collinear/reference/collinear.html) to select a non-redundant subset by prioritizing predictors by their predictive power for the response.

```{r}
trees_selection <- collinear::collinear(
  df = trees,
  responses = "log_trees",
  predictors = trees_extremes,
  f = collinear::f_numeric_rf,
  max_vif = 2.5,
  quiet = TRUE
)
```

The selected predictors for `trees` are:

```{r}
trees_selection$log_trees$selection
```

The function also returns a linear formula ready for use, and a clean dataframe ready for analysis.

```{r}
trees_selection$log_trees$formulas$linear
```

### Conditional Inference Tree

With a clean predictor set, we can fit an interpretable regression tree.

A conditional inference tree provides an interpretable summary of how the selected predictors interact to explain tree species richness. The model is fitted with [`partykit::ctree()`](https://cran.r-project.org/package=partykit), which uses permutation tests to select splits without overfitting bias.

```{r}
trees_model <- partykit::ctree(
  formula = trees_selection$log_trees$formulas$linear,
  data = trees |> 
    sf::st_drop_geometry() |> 
    na.omit(),
  control = partykit::ctree_control(
    minbucket = 100 #minimum cases in a terminal leaf
  )
)
```

The conditional inference tree below shows the sequence of environmental thresholds that best explain differences in tree species richness. Each internal node shows the splitting variable, threshold, and p-value of the permutation test; each terminal node (leaf) shows a boxplot with the richness values of the samples it contains.

```{r, fig.width = 15, fig.height = 10}
plot(x = trees_model)
```
**Interpretation examples**

The results of conditional inference trees are easy to interpret when the model has a reasonable number of terminal nodes. Below I discuss the climate conditions leading to the highest and lowest predicted richness according to the model.

**Node 15**, at the right, has the highest predicted median richness. The relevant climate thresholds leading to maximum tree richness are:

+ Minimum air humidity higher than 53%.
+ Rainfall of the wettest month higher than 115mm.
+ Rainfall of the wettest month higher than 451mm.

**Node 7**, third from the left, shows the lowest predicted tree richness, and features these climate thresholds:

+ Minimum air humidity smaller than 53%.
+ Rainfall of the wettest month higher than 104mm.
+ Maximum cloud cover higher than 36%.


### Prediction Map

Let's map the model predictions back onto the hexagonal grid.

Assigning predictions back to the grid reveals how well this simple model captures geographic variation in richness. Notice that the prediction of `partykit::ctree()` can be expressed as a quantile of the cases within each terminal node. Below we show the prediction for the quantile 0.75.

```{r, message = FALSE}
trees$predicted <- predict(
  object = trees_model,
  newdata = sf::st_drop_geometry(trees),
  type = "quantile",
  at = 0.75 #prediction at quantile 0.75
) |> 
  exp() |> #convert log(trees) to natural range
  as.integer() |> 
  as.factor()

mapview::mapview(
  trees,
  zcol = "predicted",
  layer.name = "Predicted richness",
    col.regions = viridis::turbo(
    n = 8,
    direction = 1
  ),
  color = NULL
)
```

## Conclusion

This example walked through extreme-climate predictor selection, a conditional inference tree, and spatial prediction onto a hexagonal grid — a simple but interpretable pipeline for exploring tree richness gradients.

Want to go further? Try a random forest or a spatial model, or compare richness patterns between Mesoamerican and non-Mesoamerican grid cells. Happy modelling!
