---
title: "trees"
output:
  rmarkdown::html_document:
    toc: true
    toc_title: "Contents"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval      = TRUE,
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.align = "center",
  dpi = 150,
  out.width = "100%"
)
```


## Overview

The `trees` dataset focuses on the distribution of the tree species present in Mesoamerica across North and South America. The presence of Mesoamerican trees was obtained from the **Tree Biodiversity Network (BIOTREE-NET)** dataset (project now defunct). The distribution of these species across the Americas was derived from **GBIF**. The data represents *richness of Mesoamerican trees in the Americas* for 3,373 hexagonal grid cells across the Americas, combined with 50 environmental predictor variables from different sources. The dataset was originally compiled for Benito et al. (2013).

Its main features are:

- *Americas scope*: Hexagonal cells covering longitudes -125.3° to -34.3° and latitudes -34.4° to 49.9°.
- *Single response*: `trees`, an integer count of tree species richness per hexagonal cell.
- *Rich environmental predictors*: 50 predictors spanning 10 categories (climate, soil, vegetation, geography, and more).

We designed it to support regression modelling, multicollinearity filtering, and spatial analysis of tree diversity patterns.

## Setup

```{r, include = FALSE}
if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak")
}
library(pak)
pak::pkg_install(
  c(
    "dplyr",
    "sf",
    "mapview",
    "partykit",
    "viridis",
    "collinear",
    "spatialData"
  ),
  ask = FALSE
)
```

The following R libraries and example data are required to run this tutorial.

```{r, message = FALSE, warning = FALSE}
library(dplyr)
library(sf)
library(mapview)
library(partykit)
library(viridis)
library(collinear)
library(spatialData)

data(
  trees,
  trees_response,
  trees_predictors,
  package = "spatialData"
)

trees_colors <- grDevices::hcl.colors(
  n = 7,
  palette = "Hawaii",
  rev = TRUE
)
```

## Description

The dataset is an `sf` data frame with `r nrow(trees)` rows and `r ncol(trees)` columns, and 989 cells with `NA` in the `trees` response variable (cells where environmental data is valid but no tree species were found in the relevant databases). The first 10 records and all columns but `geometry` are shown below.

```{r}
trees |>
  head(n = 10) |>
  dplyr::glimpse()
```

The hexagonal grid cells cover the Americas between approximately 50°N and 35°S. The map below shows the grid coloured by tree richness.

```{r, warning = FALSE}
mapview::mapview(
  trees,
  zcol = "trees",
  layer.name = "trees",
  col.regions = trees_colors,
  color = NULL,
  at = c(1, 10, 100, 1000, 5000, 10000, 20000)
)
```

## Response Variable

The response variable `trees` (named in the vector `trees_response`) is an **integer count** of tree species richness per hexagonal cell, derived from GBIF and BIOTREE-NET occurrence records acquired *circa* 2013.

```{r}
summary(trees$trees)
```


The dataset includes 50 numeric predictors, named in the vector `trees_predictors`.

```{r}
trees_predictors
```

Please check the documentation of [trees](https://blasbenito.github.io/spatialData/reference/trees.html) to find out their descriptions.

## Raw Data

You can download the full raw presence data with `trees_extra()`.

```{r}
trees_presence <- spatialData::trees_extra()
head(trees_presence)
```


## Example Usage

In this example we use `trees` to assess how climate extremes drive tree richness across the Americas.

- *Log Transformation*: to facilitate the visualization of model results.
- *Variable Selection*: multicollinearity analysis of the selected environmental variables.
- *Conditional Inference Trees*: fits a permutation-based tree model to explain tree richness patterns.
- *Prediction Map*: maps the model predictions back to the hexagonal grid.

### Log Transformation

Before proceeding with the variable selection, we log-transform the response variable `trees` into a new column `log_trees` to facilitate the visualization of model results. 

Tree species richness spans several orders of magnitude across the Americas, producing a strongly skewed distribution that makes terminal node summaries in the conditional inference tree difficult to read. While log-transforming count data is no longer recommended for linear models, it has no meaningful effect on split selection in `ctree()`, which relies on rank-based permutation tests that are invariant to monotone transformations of the response.

```{r}
trees <- trees |> 
  dplyr::mutate(
    log_trees = log(trees)
  )
```

### Variable Selection

Here we want to focus on the predictors indicating extreme environmental values. These are identified with the suffixes `_min` and `_max` in `trees_predictors`, with the exception of rainfall, where `rainfall_driest_month` and `rainfall_wettest_month` represent the monthly minimum and maximum respectively. The code below selects these predictors, and excludes `ndvi_*`.


```{r}
trees_extremes <- grep(
  pattern = "^(?!.*ndvi).*(min|max|driest_month|wettest_month)",
  x = colnames(trees),
  perl = TRUE,
  value = TRUE
)

trees_extremes
```

We apply [`collinear::collinear()`](https://blasbenito.github.io/collinear/reference/collinear.html) to select a non-redundant subset of predictors by prioritizing predictors by their predictive power for the response.

```{r}
trees_selection <- collinear::collinear(
  df = trees,
  responses = "log_trees",
  predictors = trees_extremes,
  f = collinear::f_numeric_rf,
  max_vif = 2.5,
  quiet = FALSE
)
```

The selected predictors for `trees` are:

```{r}
trees_selection$log_trees$selection
```

The function also returns a linear formula ready for use, and a clean dataframe ready for analysis.

```{r}
trees_selection$log_trees$formulas$linear
```

### Conditional Inference Tree

Conditional inference trees provide an interpretable summary of how the selected predictors interact to explain tree species richness. The model is fitted with [`partykit::ctree()`](https://cran.r-project.org/package=partykit), which uses permutation tests to select splits without overfitting bias.

```{r}
trees_model <- partykit::ctree(
  formula = trees_selection$log_trees$formulas$linear,
  data = trees |> 
    sf::st_drop_geometry() |> 
    na.omit(),
  control = partykit::ctree_control(
    minbucket = 100 #minimum cases in a terminal leaf
  )
)
```

The conditional inference tree below shows the sequence of environmental thresholds that best explain differences in tree species richness. Each internal node shows the splitting variable, threshold, and p-value of the permutation test; each terminal node (leaf) shows a boxplot describing the distribution of tree richness of the samples it contains.

```{r, fig.width = 15, fig.height = 10}
plot(x = trees_model)
```
The results of conditional inference trees are easy to interpret when the model has a reasonable number of terminal nodes. This is the reason behind our aggressive predictor selection above!

For example, **Node 15**, at the right, has the highest predicted median richness. The relevant climate thresholds leading to maximum tree richness are:

+ Minimum air humidity higher than 53%.
+ Rainfall of the wettest month higher than 114mm.
+ Rainfall of the wettest month higher than 451mm.

This is the subset of climate extremes leading to a maximum richness in this dataset.

On the other hand, **Node 7**, third from the left, shows the lowest predicted tree richness, and features these climate thresholds:

+ Minimum air humidity smaller than 53%.
+ Rainfall of the wettest month lower than 104mm.
+ Maximum cloud cover higher than 36%.

Let's map the model predictions back onto the hexagonal grid to see how these combinations of environmental conditions look in space.

### Prediction Map

The code below predicts and maps the conditional inference tree. 

Notice that the prediction of `partykit::ctree()` can be expressed as a quantile of the cases within each terminal node, and below we generate the prediction for the quantile 0.75. This number isn't better than others, but it's selected here to show the capabilities of conditional inference trees.

```{r, message = FALSE, warning = FALSE}
trees$predicted <- predict(
  object = trees_model,
  newdata = sf::st_drop_geometry(trees),
  type = "quantile",
  at = 0.75 #prediction at quantile 0.75
) |>
  exp() |> #convert log(trees) to natural range
  as.integer() |>
  as.factor()

mapview::mapview(
  trees,
  zcol = "predicted",
  layer.name = "Predicted richness",
  col.regions = trees_colors,
  color = NULL,
  at = as.numeric(levels(trees$predicted))
)
```

Notice that our prediction filled all NA cases in `trees$trees`!

## Conclusion

This example uses the `trees` dataset to explore how a conditional inference tree can  be fitted with extreme climate predictors to explain and predict Mesoamerican tree richness across the Americas. It is a simple yet interpretable model useful to assess richness gradients, but there many other approaches to test with this rich dataset.

## References

Benito, B.M., Cayuela, L., & Albuquerque, F.S. (2013). The impact of modelling choices in the predictive performance of richness maps derived from species-distribution models: Guidelines to build better diversity models. *Methods in Ecology and Evolution*, 4(4), 327–335. https://doi.org/10.1111/2041-210X.12022

Cayuela, L., Gálvez-Bravo, L., Pérez Pérez, R., de Albuquerque, F.S., Golicher, D.J., Zahawi, R.A., et al. (2012). The Tree Biodiversity Network (BIOTREE-NET): prospects for biodiversity research and conservation in the Neotropics. *Biodiversity & Ecology*, 4, 211–224. https://doi.org/10.7809/b-e.00078


