---
title: "vi"
output:
  rmarkdown::html_document:
    toc: true
    toc_title: "Contents"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval      = TRUE,
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.align = "center",
  dpi = 150,
  out.width = "100%"
)
```


## Overview

The `vi` dataset captures the variation of the Normalized Difference Vegetation Index (NDVI) at global scale, with ~10 000 records spanning 6 continents and 12 biomes, and a large set of numeric and categorical predictors (see `vi_predictors`) with consistent naming patterns (e.g., `temperature_*`, `rainfall_*`, `swi_*`).

Its main features are:

- *Complete case study* - No missing values, clean data ready for analysis.
- *Global scope* - 6 continents, all major climate zones, 12 biomes.
- *Grouped/stratified analysis* - Multiple hierarchical groupings available.

This dataset has several versions of the same response (see [vi_responses]) to support a wide range of analytical approaches: linear, count and ordinal regression, binary and multiclass classification, and mixed-effects models. 

Its large number of predictors allows testing dimensionality reduction and multicollinearity filtering methods, while its dense point cloud helps test thinning, interpolation, spatial cross-validation, and spatial autocorrelation analyses.

## Setup

The following R libraries are required to run this tutorial:

```{r, include = FALSE}
if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak")
}
library(pak)
pak::pkg_install(
  c(
    "tidyr",
    "dplyr",
    "sf",
    "mapview",
    "ggplot2",
    "mgcv",
    "plotmo",
    "spatialRF",
    "collinear",
    "spatialData"
  ),
  ask = FALSE
)
```

```{r, message = FALSE, warning = FALSE}
library(tidyr)
library(dplyr)
library(sf)
library(mapview)
library(ggplot2)
library(mgcv)
library(plotmo)
library(spatialRF)
library(collinear)
library(spatialData)

data(
  vi,
  vi_responses,
  vi_predictors,
  package = "spatialData"
  )

vi_colors <- grDevices::hcl.colors(
  n = 5,
  palette = "Fall",
  rev = TRUE
)
```


## Data Structure

The dataset is an `sf` data frame with `r nrow(vi)` rows and `r ncol(vi)` columns, and no missing data. The first 10 records and all columns but `geometry` are shown below.

```{r}
vi |>
  sf::st_drop_geometry() |>
  head(n = 10) |>
  dplyr::glimpse()
```

A larger version of the dataset (~30k records) with data for non-vegetated regions can be loaded via `vi_extra()`.

```{r, eval = FALSE}
vi_large <- spatialData::vi_extra()
nrow(vi_large)
```

The difference between `vi` and its large version is that the records in `vi` are limited to vegetated areas (`vi$vi_numeric > 0`).

There is also a smaller version of the dataset, `vi_smol`, ideal for testing ideas before scaling up to the full dataset.

```{r}
data(vi_smol)
nrow(vi_smol)
```


The map below shows `vi` and its main response variable, `vi_numeric`.


```{r, warning = FALSE}
mapview::mapview(
  vi,
  zcol = "vi_numeric",
  layer.name = "vi_numeric",
  col.regions = vi_colors
)
```

There are **5 different encodings** (see `vi_responses`) of the same underlying vegetation index in `vi`, enabling different modeling approaches:

| Variable | Type | Description | Use Case |
|----------|------|-------------|----------|
| `vi_numeric` | Continuous (0-1) | Raw NDVI values | Linear regression, GAM |
| `vi_counts` | Integer | `vi_numeric * 1000` as integer | Poisson/negative binomial regression |
| `vi_binomial` | Binary (0/1) | Thresholded to 0 or 1 | Logistic regression, classification |
| `vi_categorical` | Character | 5 categories based on quantiles | Multinomial models |
| `vi_factor` | Factor | `vi_categorical` as ordered factor | Ordinal regression |

The dataset also includes **58 predictor variables from 11 datasources**, of which 47 are numeric and 11 are categorical.

```{r}
vi_predictors
```
The data in `vi` provides a comprehensive insight into the relationship between soil, climate, and plant health and productivity across the world's biomes. A good example is the plot below, which shows how NDVI is limited by maximum soil temperature and the minimum soil water index in similar ways across very different biomes.

```{r, fig.width=8, fig.height=10, warning = FALSE}
vi |> 
  ggplot2::ggplot() +
  ggplot2::aes(
    x = soil_temperature_max,
    y = swi_min,
    color = vi_numeric
  ) + 
  ggplot2::facet_wrap(
    facets = ~biogeo_biome, 
    ncol = 2,
    scales = "free"
    ) + 
  ggplot2::geom_point() + 
  ggplot2::scale_color_continuous(
    palette = vi_colors
  ) + 
  ggplot2::labs(
    x = "Maximum Land Surface Temperature (Cº)",
    y = "Minimum Soil Water Index (%)",
    color = "NDVI"
  ) +
  ggplot2::theme_bw()
```

## Example Usage

This brief example shows how `vi` can be used to fit a spatially-aware Generalized Additive Model (GAM) that explicitly accounts for spatial autocorrelation in global NDVI data.

### Spatial Components

The code below computes the distance matrix between locations in `vi` we need to assess the spatial auto-correlation (SAC) in the model residuals.

```{r}
vi_distance_matrix_meters <- vi |> 
  sf::st_distance() |> 
  unclass() #removes units (meters)

vi_distance_matrix_km <- vi_distance_matrix_meters / 1000
rm(vi_distance_matrix_meters)
```


We also need the geographic coordinates of the cases, which are encoded as well-known text in the `geometry` column.

```{r}
vi_coordinates <- vi |> 
  sf::st_coordinates() |> 
  as.data.frame() |> 
  dplyr::rename(
    latitude = Y,
    longitude = X
  )

vi <- dplyr::bind_cols(
  vi,
  vi_coordinates
)
```


### Multicollinearity Filtering

The `vi_predictors` vector contains both numeric and categorical variables. To simplify our modelling approach we will focus on numeric predictors only, which can be extracted using `collinear::identify_numeric_variables()`.


```{r}
vi_predictors_numeric <- collinear::identify_numeric_variables(
    df = sf::st_drop_geometry(vi),
    predictors = vi_predictors
  )$valid

vi_predictors_numeric
```

Many predictors in `vi_predictors_numeric` are highly correlated (e.g., `temperature_mean`, `temperature_max`, `temperature_min`). The [`collinear::collinear()`](https://blasbenito.github.io/collinear/reference/collinear.html) function selects a non-redundant subset from a ranking based on univariate GAM models.

```{r}
vi_selection <- collinear::collinear(
  df = vi,
  responses = "vi_numeric",
  predictors = vi_predictors_numeric,
  f = collinear::f_numeric_gam,
  quiet = FALSE
)
```

The names of the selected predictors are stored in this vector.

```{r}
vi_selection$vi_numeric$selection
```
The function also returns a GAM formula ready for use.

```{r}
vi_formula <- vi_selection$vi_numeric$formulas$smooth
vi_formula
```

### Non-spatial GAM Model

Let's use the formula above to fit a regular GAM model with environmental predictors only.


```{r}
vi_model_non_spatial <- mgcv::gam(
  formula = vi_formula,
  data = sf::st_drop_geometry(vi),
  select = TRUE
)

summary(vi_model_non_spatial)
```

The model looks impressive, but let's take a look at the SAC of the model residuals.

```{r, fig.width=6, fig.height=3}
vi_model_non_spatial_sac <- spatialRF::moran_multithreshold(
  x = stats::residuals(vi_model_non_spatial),
  distance.matrix = vi_distance_matrix_km,
  distance.thresholds = c(100, 500, 1000, 2500, 5000)
)
```

The SAC in this model's residuals is quite high! This indicates that the model is achieving a good fit partly by exploiting spatial proximity rather than genuine environmental signal. SAC is likely compressing standard errors and pushing p-values down, making smooth terms look more significant than they truly are.

Let's take a look at an approach that might help mitigate this problem.

### Spatially-aware GAM

The idea here is to provide the GAM Model with a predictor representing spatial proximity between records. This predictor will absorb part of the overall explained variance, resulting in more honest effect estimates for the environmental predictors.
 
For this approach we add the term `s(latitude, longitude, bs = "sos")` to the original model formula. This is a smooth-on-a-sphere interaction of the original latitudes and longitudes in degrees. 

This term needs also an argument `k` (number of knots) with a value high enough to give the GAM the flexibility required to match the autocorrelation structure of the data. In the case of `vi`, with a width of about 40000km, and a height of about 12000km (check `sf::st_bbox(vi)`), and a SAC still significant at 1000km, we can think of an imaginary grid with a cell size of about 1000x1000km, with `40 * 12` cells for a total of 480.
 
To compose the new formula, first we need to convert the original formula to character (`paste(deparse(...))`), and then add the new term.

```{r}
vi_formula_spatial <- formula(
  paste(
    paste(deparse(vi_formula), collapse  = " "),
    "+ s(latitude, longitude, bs = 'sos', k = 480)"
  )
)

vi_formula_spatial
```

Let's take a look at the new model!

```{r}
vi_model_spatial <- mgcv::gam(
  formula = vi_formula_spatial, 
  data = sf::st_drop_geometry(vi),
  select = TRUE
  )

summary(vi_model_spatial)
```

Notice that individual smooth terms explain less variance than in the non-spatial model. The `s(latitude, longitude)` term is absorbing part of the geographic structure that the environmental predictors were capturing before, so this is expected. The goal here is not to maximize R², but to get effect estimates that reflect genuine environmental signal rather than spatial proximity.

```{r, fig.width=6, fig.height=3}
vi_model_spatial_sac <- spatialRF::moran_multithreshold(
  x = stats::residuals(vi_model_spatial),
  distance.matrix = vi_distance_matrix_km,
  distance.thresholds = c(100, 500, 1000, 2500, 5000)
)
```

SAC in the residuals is much lower now, indicating that the spatial term of our model is doing a great job!

Now, if you are curious about how the spatial term looks like, the map below shows its contribution to predicted NDVI with the environmental predictors factored out. 

```{r, warning = FALSE}
vi$spatial_effect_sos <- mgcv::predict.gam(
  object = vi_model_spatial,
  type = "terms",
  terms = "s(latitude,longitude)"
)[, 1]

mapview::mapview(
  vi,
  zcol = "spatial_effect_sos",
  layer.name = "Spatial component",
  col.regions = grDevices::hcl.colors(
    n = 10,
    palette = "Green-Brown",
    rev = TRUE
  )
)
```

Positive values (teal) mark areas where vegetation is more productive than local conditions alone would predict, while negative values (brown) mark the opposite. The likely causes are things the predictors don't measure, such as land use history, soil legacy effects, or plant adaptations to specific climates.


### Model Fit

The code below predicts both models on `vi` to plot observed vs predicted NDVI. 

```{r, fig.width=8, fig.height=4}
vi_long <- vi |>
  dplyr::mutate(
    non_spatial = stats::fitted(vi_model_non_spatial),
    spatial = stats::fitted(vi_model_spatial)
  ) |>
  tidyr::pivot_longer(
    cols = c("non_spatial", "spatial"),
    names_to = "model",
    values_to = "prediction"
  )

ggplot2::ggplot(vi_long) +
  ggplot2::aes(
    x = vi_numeric,
    y = prediction,
    color = vi_numeric
  ) +
  ggplot2::facet_wrap(
    facets = ~model,
    ncol = 2
  ) +
  ggplot2::geom_point(
    alpha = 0.5,
    size = 0.5
  ) +
  ggplot2::scale_color_continuous(
    palette = vi_colors
  ) +
  ggplot2::geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed"
  ) +
  ggplot2::labs(
    title = "Observations vs. Predictions",
    x = "Observed NDVI",
    y = "Predicted NDVI"
  ) +
  ggplot2::coord_equal() +
  ggplot2::theme_bw() + 
  ggplot2::theme(legend.position = "none")
```
Both models do a pretty good job at matching predictions with observations, but the spatial model shows a narrower band tightly grouped around the diagonal.

## Conclusions

The example above uses `vi` to demonstrate how how incorporating the geographic structure of the data along with the environmental predictors results in a better and more realistic model. But this is only the beginning! `vi` is large and rich enough to support many other modelling approaches, such as structural equation modelling, spatial clustering, mixed modelling, and many others. 

Enjoy it!

